A DPG method for compressible flow problems

### 2012-06-25 docs/Jesse/Proposal/abstract.tex
adding stuff to tex files of jesse proposal

\author{Jesse Chan\\Committee: \vspace{-0.3cm}\\Dr. Leszek Demkowicz (co-supervisor), Dr. Robert Moser (co-supervisor),\vspace{-0.3cm}\\ Dr. Tom Hughes, Committee Member 2, Committee Member 3}

@@

Residual-based stabilizations can alternatively be viewed as modifying the standard finite element test space, and consequently the norm in which the finite element method converges. Demkowicz and Gopalakrishnan generalized this idea in 2009 by introducing the discontinous Petrov-Galerkin (DPG) method with optimal test functions, where test functions are determined such that they minimize the discrete linear residual in a dual space.  Under the ultra-weak variational formulation, these test functions can be computed locally to yield a symmetric, positive-definite system.  The DPG method was quickly used to solve the convection and convection-diffusion problems \cite{DPG1,DPG2,DPG3}, and has recently been shown to yield a robust higher order adaptive method for the convection-diffusion model problem in arbitrary dimensions \cite{DPGrobustness,DPGrobustness2}.

@@

A large thrust of the numerical work here has focused on the development of a 2D compressible flow code under the parallel DPG Camellia codebase, developed by Nathan Roberts at ICES \cite{Camellia}.  In particular, we have developed a framework allowing for rapid implementation of problems and the easy application of higher order and $hp$-adaptive schemes based on a natural error representation function that stems from the DPG residual \cite{DPG2, DPG3}. Additionally, the effectiveness and necessity of anisotropic refinement schemes for problems in CFD has been demonstrated several times in the literature \cite{anisotropy1,anisotropy2}. The DPG error representation function has been shown to yield an effective and natural vector-valued residual with which to drive refinement, which we hope to generalize to anisotropic adaptive schemes.

@@

I will apply the DPG method to a several convection diffusion problems which mimic difficult problems in compressible flow simulations, including boundary layer and closed streamline problems.  I will also solve a viscous Burgers' equation as an extension of DPG to nonlinear problems.  Finally, the effectiveness of DPG as a numerical method for compressible flow will be assessed with the application of DPG to common transonic and superonic benchmark problems.  In particular, I will use DPG to solve the flat plate problem, the ramp problem, flow over a Gaussian bump, and flow over an airfoil over a range of Mach numbers and laminar Reynolds numbers.

Time permitting, I will also explore the regularization of the inviscid convection and Euler equations using the full convection-diffusion and Navier-Stokes equations in the small-viscosity limit.

>>>>

\author{Jesse Chan\\Committee: \vspace{-0.3cm}\\Dr. Leszek Demkowicz (co-supervisor), Dr. Robert Moser (co-supervisor),\vspace{-0.3cm}\\ Dr. Todd Arbogast, Dr. Omar Ghattas, Dr. Venkat Raman}

@@

Residual-based stabilizations can alternatively be viewed as modifying the standard finite element test space, and consequently the norm in which the finite element method converges. Demkowicz and Gopalakrishnan generalized this idea in 2009 by introducing the Discontinous Petrov-Galerkin (DPG) method with optimal test functions, where test functions are determined such that they minimize the discrete linear residual in a dual space.  Under the ultra-weak variational formulation, these test functions can be computed locally to yield a symmetric, positive-definite system.  The DPG method was quickly used to solve the convection and convection-diffusion problems \cite{DPG1,DPG2,DPG3}, and has recently been shown to yield a robust higher order adaptive method for the convection-diffusion model problem in arbitrary dimensions \cite{DPGrobustness,DPGrobustness2}.

@@

A large thrust of the numerical work here has focused on the development of a 2D compressible flow code under the parallel DPG Camellia codebase, developed by Nathan Roberts at ICES \cite{Camellia}.  In particular, we have developed a framework allowing for rapid implementation of problems and the easy application of higher order and $hp$-adaptive schemes based on a natural error representation function that stems from the DPG residual \cite{DPG2, DPG3}. Additionally, the effectiveness and necessity of anisotropic refinement schemes for problems in CFD has been demonstrated several times in the literature \cite{anisotropy1,anisotropy2}. The DPG error representation function has been shown to yield an effective and natural residual with which to drive refinement, which we hope to generalize to anisotropic adaptive schemes.

@@

I will apply the DPG method to a several convection diffusion problems which mimic difficult problems in compressible flow simulations, including boundary layer and closed streamline problems.  I will also solve a viscous Burgers' equation as an extension of DPG to nonlinear problems.  Finally, the effectiveness of DPG as a numerical method for compressible flow will be assessed with the application of DPG to common transonic and superonic benchmark problems.  In particular, I will use DPG to solve the flat plate problem, the ramp problem, and flow over a Gaussian bump over a range of Mach numbers and laminar Reynolds numbers.

Time permitting, I will also attempt to simulate flow over an airfoil, and to explore the regularization of the inviscid convection and Euler equations using the full convection-diffusion and Navier-Stokes equations in the small-viscosity limit.

### 2012-06-25 docs/Jesse/Proposal/outline.tex
committing changes to proposal tex/bib files
adding stuff to tex files of jesse proposal

\title{Proposal outline/notes for DPG and Compressible Navier-Stokes}

@@

\committeemembers
	[Thomas J. R. Hughes]
	[Somebody]
	[Somebody]
	{Eh}

@@

\include{proposalIntro}

@@

\include{DPGtheory}

@@

\begin{itemize}
\item Describe mathematically the issue with standard Galerkin.
\item Give proofs of robustness and numerical results (from paper)\cite{ChanHeuerBui-ThanhDemkowicz12}
\end{itemize}
\todo{Transition between paper and this bit}

\include{DPGrobustness}

@@

\include{NonlinDPG}
\include{BurgersNotes}
\include{NSNotes}
\include{areaReq}

@@

%\chapter{Appendix}
\appendices
\input{paper_appendix.tex}
\input{hyperbolicDPG.tex}

>>>>

\title{A DPG method for compressible flow problems}

@@

\committeemembers
	[Committee member 1]
	[Committee member 2]
	{Committee member 3}

@@

\input{proposalIntro}

\chapter{Range of problems}

\input{problemRange}

@@

\input{DPGtheory}

@@

\input{DPGrobustness}

@@

\input{NonlinDPG}
\input{BurgersNotes}
\input{NSNotes}
\input{areaReq}

@@

%\chapter{Appendix}
\appendix
\input{paper_appendix.tex}
%\input{hyperbolicDPG.tex}

### 2012-06-25 docs/Jesse/Proposal/problemRange.tex
adding stuff to tex files of jesse proposal

\todo{write blurb on how comp NS is used - small Re limit}
\todo{write blurb on turbulence}

@@

Under appropriate assumptions on the behavior of density and temperature, the behavior of the compressible Navier-Stokes equations can be sufficiently represented by the incompressible Navier-Stokes equations for low speed laminar fluid flows.

For viscous ``creeping'' flows, the incompressible Navier-Stokes equations reduce down to the Stokes equations. \cite{stokesDPG}

@@

The scalar convection-diffusion problem is the prototypical model problem for solving the full Navier-Stokes equations; most stabilized methods consider first the scalar convection-diffusion equation as a test case before attempting a solution of the full Navier-Stokes equations. As discussed previously, an important feature of the convection-diffusion equation is that solutions can develop  boundary layers, a physical feature found in most applications of interest for compressible flow.

The scalar convection-diffusion equation is also mathematically important

\todo{add perspective on convection-diffusion}

>>>>

As mentioned before, the compressible Navier-Stokes equations are especially of interest in the simulation of high-speed air flows. In other contexts, however, the compressible Navier-Stokes equations may be simplified based on physical assumptions about the problem at hand. We briefly cover several simplifying assumptions common in CFD applications.

@@

Under appropriate assumptions on the behavior of density and temperature, the behavior of the compressible Navier-Stokes equations can be sufficiently represented by the incompressible Navier-Stokes equations for low speed laminar fluid flows. The study of the incompressible Navier-Stokes equations is an open area in mathematics, and is one of the most famous Millenium Problems posed by the Clay Mathematics Institute. The equations of incompressible flow pose a difficult problem computationally as well, in part due to the problem of the simulation of turbulent phenomena.

For highly viscous ``creeping'' flows, the incompressible Navier-Stokes equations reduce down to the Stokes equations. We remark that determining good finite element spaces for the Stokes problem is still an active area of research. \cite{stokesFEM} lists several choices of finite element discretizations suitable for the Stokes equation.

The scope of this dissertation will not deal with these two equations --- the Stokes equations are treated in \cite{stokesDPG}, and the incompressible Navier-Stokes are covered in an upcoming dissertation.

@@

The scalar convection-diffusion equation models mathematically the distribution of the concentration $u$ of a substance in a medium due to both convective and diffusive effects. The general form of the equation in $\mathbb{R}^n$ is 
\[
\beta\cdot \grad u - \epsilon \Delta u = f,
\]
where $\beta \in \mathbb{R}^n$ is a vector specifying the direction of convection.

Scalar convection-diffusion has significant historical importance, as it is the prototypical model problem for solving the full Navier-Stokes equations --- most stabilized methods consider first the scalar convection-diffusion equation as a test case before attempting a solution of the full Navier-Stokes equations. As discussed previously, an important feature of the convection-diffusion equation is that solutions can develop boundary layers whose thickness depends on the viscosity, a physical feature found in most applications of interest for compressible flow.
